Speeding up eQTL scans in the BXD population using GPU

### 2019-10-02 paper.tex
added some reference, added a paragraph about why use GPU

\keywords{Linear Model \\ Genome Scan \\ GPU}

@@

@@

\correspondingauthoraffiliation{1}{Corresponding author: Saunak Sen, Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. sen@uthsc.edu}
\vspace{-34pt}% Only used for adjusting extra space in the left column of the first page

\noindent The BXD mouse strains are an important set of recombinant
inbred lines with deep phenotypic characterization.  Since the strains
have been genotyped and sequenced, once a set of strains is
phenotyped, the data can be immediately used for QTL (quantitative
trait locus) mapping, and associational analysis with
previously-collected phenotypes.  For ``omic'' phenotypes collected
using high-throughput technologies, additional analyses (such as
transcriptional network construction or causal mediation analyses) are
possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
seeks to facilitate this process by providing a searchable database of
phenotypes and genotypes in a variety of organisms (including mouse,
rat, and Arabidopsis).  It provides a suite of interactive tools for
browsing the data, QTL mapping, correlational analyses, network
construction, and genome browsing.  Our goal is to develop a real-time
backend for GeneNetwork to perform eQTL analysis (where we perform
single-trait QTL mapping for thousands of ``omic'' traits) for
important genetic reference populations such as the BXD population.

%% Computational demands in omics are growing exponentially. For
%% instance, eQTL analysis requires processing genomewide expression and
%% marker data; this can take minutes, hours or days depending on problem
%% size.  Scientists want to interact with the results and real-time
%% processing would be valuable.  Our goal is to build a backend for
%% GeneNetwork that will enable real-time genome scans for important
%% genetic reference populations such as the BXD population.

To perform eQTL scans in the BXD population, one has to simply perform
as many genome scans as there are phenotypes.  This can be done in an
``embarrassingly parallel'' fashion by using standard algorithms for
QTL analysis, such as those employed by R/qtl (cite).  In practice,
this is too slow, and substantial speedups are possible.  [Mention
  ideas that have been used before] For example, by using the
Haley-Knott algorithms using genotype probabilities on batches of
phenotypes with the same missing data pattern, instead of using the EM
algorithm on each phenotype individually, substantial speedups are
possible.  This is a well-known trick and is used by R/qtl.
Additionally, if only additive effects are tested, or if the
population has only two genotype categories (as in a backcross or
recombinant inbred line), then matrix multiplication can be used to
perform Haley-Knott regression \citep{shabalin2012matrix}.

The matrix multiplication with large matrices is the most computationally intensive and time-consuming 
to perform eQTL analysis.  By dividing large matrices into small block matrices, 
one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, 
Intel Kernel Math Library (KML) in Revolution R/Matlab, or even GPU CUDA in Matlab.  
For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it (cite : Matrix eQTL).

@@

	\caption{Schematic of data and correlation calculation: $Y$ is
		the expression phenotype matrix, $G$ is the matrix of
		genotypes, and $R=Y^{\prime}G$ is the matrix of
		correlations.  The LOD scores are a function of the
		correlation matrix.}
	\label{MatrixMult}
	\includegraphics[scale = .4]{figs/YGmatrix.png}
\end{figure}

>>>>

\keywords{Linear Model \\ Genome Scan \\ BXD \\ GPU}

@@

% This paper aims to answer three questions :
% 1: Why BXD : reference population
% 2: Why Speed up : GN backend and benefit other researchers
% 3: Why GPU: Growth of CPU computing is slowing down, GPU as a new computing platform is competitive, low cost per core, and faster.

% Paper Layout:
% Abstract 
% Intro: 
%P1: Why BXD; 
%P2: Why Speed Up; 
%P3: Why GPU 
%P4: Other works. 
%P5: Outline of Paper

% Method: 
%P1: Intro to LM and Acceleration techniques; 
%P2: Data Layout; 
%P3: LM; 
%P4: Covariates
%P4 Acceleration Techniques;

% Result: 
%P1: Linear Model; 
%P2: Result from using different acceleration techniques; 
%P3: Computing platform. 
%P4: Software Availability; 
%P5: Data Availability

% Conclusion: 
%P1: Our final result from acceleration. 
%P2: Limitations; 
%P3: Future work.

@@

\correspondingauthoraffiliation{1}{Corresponding author: Saunak Sen, Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis, TN. sen@uthsc.edu}
\vspace{-23pt}% Only used for adjusting extra space in the left column of the first page

%Why BXD
\noindent The BXD mouse strains are an important set of recombinant
inbred lines with deep phenotypic characterization.  Since the strains
have been genotyped and sequenced, once a set of strains is
phenotyped, the data can be immediately used for quantitative
trait locus (QTL) mapping, and associational analysis with
previously-collected phenotypes.  For ``omic'' phenotypes collected
using high-throughput technologies, additional analyses (such as
transcriptional network construction or causal mediation analyses) are
possible.

%Why Speed Up
The GeneNetwork project (\url{http://gn2.genenetwork.org})
seeks to facilitate this process by providing a searchable database of
phenotypes and genotypes in a variety of organisms (including mouse,
rat, and Arabidopsis).  It provides a suite of interactive tools for
browsing the data, QTL mapping, correlational analyses, network
construction, and genome browsing. Our goal is to develop a real-time
backend for GeneNetwork to perform eQTL analysis (where we perform
single-trait QTL mapping for thousands of ``omic'' traits) for
important genetic reference populations such as the BXD population.

%Why GPU?
Processing large data set has been a challenge for genome scan. 
We were blessed with Moore's Law for decades, but the Central Processing Unit (CPU) is reaching to its physical limit of transistors. 
Graphics Processing Unit (GPU), originally used as an image processing component of a computer, has shown some compelling results of accelerating computation in various fields.
General Purpose computing Graphics Processor (GPGPU) became popular in the early 2000s because of its ability to natively handel matrix and vector operations. 
Such power is very attractive to the scientific computing community. 
BXD genome scan can be simplified down to matrix operations, which is quite amendable to GPU computing.

%Other Works
To perform eQTL scans in the BXD population, one has to simply perform
as many genome scans as there are phenotypes.  This can be done in an
``embarrassingly parallel'' fashion by using standard algorithms for
QTL analysis, such as those employed by R/qtl \citep{broman2003r}.  In practice,
this is too slow, and substantial speedups are possible.  [Mention
  ideas that have been used before] For example, by using the
Haley-Knott algorithms \citep{haley1992simple} using genotype probabilities on batches of
phenotypes with the same missing data pattern, instead of using the EM
algorithm \citep{lander1989mapping} on each phenotype individually, substantial speedups are
possible. This is a well-known trick and is used by R/qtl.
Additionally, if only additive effects are tested, or if the
population has only two genotype categories (as in a backcross or
recombinant inbred line), then matrix multiplication can be used to
perform Haley-Knott regression.

The matrix multiplication with large matrices is the most computationally intensive and time-consuming 
to perform eQTL analysis.  By dividing large matrices into small block matrices, 
one can achieve striking efficiency in matrix multiplication in conjunction with built-in BLAS in R, 
Intel Kernel Math Library (KML) in Revolution R/Matlab, or even GPU CUDA in Matlab.  
For instance, GOTO BLAS, only available in 64 bit R only, and KML allow the use of multiple CPU cores and provide 14-17 times 
better performance than built-in BLAS in R.  GPU-based computation shows 10 times better performance than 
the best algorithm for the CPU for single precision even if Matlab provides GPU CUDA but limited support to it  \citep{shabalin2012matrix}.

@@

	\caption{Schematic of data and correlation calculation: $Y$ is
		the expression phenotype matrix, $G$ is the matrix of
		genotypes, and $R=Y^{*\prime}G^{*}$ is the matrix of
		correlations.  The LOD scores are a function of the
		correlation matrix.}
	\label{MatrixMult}
	\includegraphics[scale = .4]{figs/YGmatrix.png}
\end{figure}
